---
title: "randombullshit2"
format: pdf
editor: visual
---

```{r}
# Install the packages
#install.packages(c("tidyverse", "nflfastR", "ggimage", "gt", "ggthemes", "ggrepel", "plotly", "vip"))
#install.packages('ggforce')
#install.packages('ggplot2')
#install.packages('ggalt')

# Load the packages
library(tidyverse)
library(nflfastR)
library(ggimage)
library(gt)
library(ggthemes)
library(ggrepel)
library(plotly)
library(vip)
library(ggforce)
library(ggplot2)
library(ggalt)
```

```{r}
#long_pbp <- load_pbp(1999:2024)
#write.csv(long_pbp, '/Users/manik/Downloads/STAT405/temp_storage/long_pbp.csv')
long_pbp <- read.csv('/Users/manik/Downloads/STAT405/temp_storage/long_pbp.csv')
```

```{r}
tbrady_passes <- long_pbp %>%
  filter(passer == 'T.Brady' & 
         play_type != 'no_play' & 
         incomplete_pass != 1 & 
         interception != 1 & 
         !is.na(air_yards) & 
         !is.na(yards_after_catch))


tbrady_passes <- tbrady_passes %>%
  arrange(desc(receiver_id))

top_receivers <- tbrady_passes %>%
  group_by(receiver_id, receiver) %>%  # Group by receiver_id and receiver
  summarize(
    total_yards_gained = sum(yards_gained, na.rm = TRUE),  # Sum yards_gained
    num_recs = n(),  # Count of plays for each receiver
    total_air_yards = sum(air_yards, na.rm = TRUE),  # Sum of air yards
    total_yards_after_catch = sum(yards_after_catch, na.rm = TRUE),  # Sum of yards after catch
    .groups = 'drop'  # Ungroup after summarizing
  )

top_receivers <- top_receivers %>%
  arrange(desc(total_yards_gained)) %>%  # Sort in descending order
  slice_head(n = 20)  # Take the top 20 rows

player_positions <- data.frame(
  receiver = c("R.Gronkowski", "J.Edelman", "W.Welker", "M.Evans", "C.Godwin", 
               "J.White", "R.Moss", "D.Amendola", "A.Hernandez", "C.Hogan", 
               "D.Branch", "B.Watson", "B.LaFell", "L.Fournette", "K.Faulk", 
               "S.Vereen", "B.Cooks", "A.Brown", "D.Woodhead", "B.Lloyd"),
  position = c("TE", "WR", "WR", "WR", "WR", 
               "RB", "WR", "WR", "TE", "WR", 
               "WR", "TE", "WR", "RB", "RB", 
               "RB", "WR", "WR", "RB", "WR")
)

# Append the positions to the top_receivers dataframe
top_receivers <- top_receivers %>%
  left_join(player_positions, by = "receiver")

top_receivers <- top_receivers %>%
  mutate(
    air_yards_per_rec = total_air_yards / num_recs,
    yards_after_catch_per_rec = total_yards_after_catch / num_recs
  )

top_receivers <- top_receivers %>%
  filter(!is.na(air_yards_per_rec) & !is.na(yards_after_catch_per_rec))

ggplot(top_receivers, aes(x = air_yards_per_rec, 
                           y = yards_after_catch_per_rec, 
                           label = receiver, 
                           color = position)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_text(vjust = -0.5, hjust = 0.5, size = 3.5) +
  geom_encircle(aes(fill = position), 
                alpha = 0.1, 
                color = NA, 
                expand = 0.08) +
  labs(
    title = "Air Yards and Yards After Catch per Reception",
    x = "Air Yards per Reception",
    y = "Yards After Catch per Reception"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

```
Interperation: Of Tom Brady's receivers, the running backs tended to have more yards after catch per reception, and the least air yards per reception. However, his tight ends and wide receivers tended to have most of their yards in the air and only a little after the catch. Some of his receivers were especially good at getting a lot of air yards.
```{r}
all_passes <- long_pbp %>% 
  filter(play_type == 'pass' &
         !is.na(passer_id)) %>%
  mutate(
    Complete = complete_pass == "1", TRUE, FALSE
  )

summary_by_passer <- all_passes %>%
  group_by(passer_id) %>%
  summarize(
    passer = first(passer),  # Get the passer corresponding to each passer_id
    num_plays = n(),
    total_air_yards = sum(air_yards, na.rm = TRUE),
    completion_percentage = mean(Complete) * 100
  )

summary_by_passer <- summary_by_passer %>%
  mutate(average_air_yards_per_play = total_air_yards / num_plays) %>%
  filter(num_plays >= 50)

ggplot(summary_by_passer, aes(x = average_air_yards_per_play, y = completion_percentage)) +
  geom_density_2d_filled() +  # Add filled 2D density contours
  labs(
    title = "2D Density Plot of Average Air Yards per Play vs. Completion Percentage",
    x = "Average Air Yards per Play",
    y = "Completion Percentage"
  ) +
  theme_minimal()
```
Most quarterbacks tend to be at 7.5 air yards pre play and a completion percentage at around 55%, and the density is pretty strongly concentrated around that point. There are a small number of quarterbacks who have a very low average air yards per play as well though. 
